---
title: "Capstone project Google Data Analytics Course R report"
author: "Oleg Kulikov"
date: "2/27/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This project uses fictional company Cyclistic to perform an analysis of historical trip data.
Data used from https://divvy-tripdata.s3.amazonaws.com/index.html is provided by Motivate International Inc
under this licence https://ride.divvybikes.com/data-license-agreement.

## Three Questions will guide the future marketing program.

1. How do annual members and casual riders use Cyclistic bikes differently?
2. Why would casual riders buy Cyclistic annual memberships?
3. How can Cyclistic use digital media to influence casual riders to become members?

You are assigned first question to answer.

## Deliverables

You will produce a report with the following deliverables:
1. A clear statement of the business task
2. A description of all data sources used
3. Documentation of any cleaning or manipulation of data
4. A summary of your analysis
5. Supporting visualizations and key findings
6. Your top three recommendations based on your analysis

## Steps of Analysis:

1. Download CVS files from the past 12 months.
2. Open each dataset in Excel to get familiar with data.
3. Uploading required packages for our data analysis and visualizations.
```{r}
library(tidyverse)
library(ggplot2)
library(janitor)
library(skimr)
```
4. Uploading excel files into RStudio Environment.

```{r Data Import}
X202101_divvy_tripdata <- read_excel("Bikedata/XLS_data/202101-divvy-tripdata.xlsx")
X202102_divvy_tripdata <- read_excel("Bikedata/XLS_data/202102-divvy-tripdata.xlsx")
X202103_divvy_tripdata <- read_excel("Bikedata/XLS_data/202103-divvy-tripdata.xlsx")
X202104_divvy_tripdata <- read_excel("Bikedata/XLS_data/202104-divvy-tripdata.xlsx")
X202105_divvy_tripdata <- read_excel("Bikedata/XLS_data/202105-divvy-tripdata.xlsx")
X202106_divvy_tripdata <- read_excel("Bikedata/XLS_data/202106-divvy-tripdata.xlsx")
X202107_divvy_tripdata <- read_excel("Bikedata/XLS_data/202107-divvy-tripdata.xlsx")
X202108_divvy_tripdata <- read_excel("Bikedata/XLS_data/202108-divvy-tripdata.xlsx")
X202109_divvy_tripdata <- read_excel("Bikedata/XLS_data/202109-divvy-tripdata.xlsx")
X202110_divvy_tripdata <- read_excel("Bikedata/XLS_data/202110-divvy-tripdata.xlsx")
X202111_divvy_tripdata <- read_excel("Bikedata/XLS_data/202111-divvy-tripdata.xlsx")
X202112_divvy_tripdata <- read_excel("Bikedata/XLS_data/202112-divvy-tripdata.xlsx")
```

## Getting Familiar with Data before merging it all together.
In example I will show just first table, but we will perform same checks on every table before merging.

```{r check the column names}
colnames(X202101_divvy_tripdata)

View(X202101_divvy_tripdata)
```

Inspecting elements and data formats to make sure that everything matches before integration.

```{r compare_df_cols}
compare_df_cols(X202101_divvy_tripdata,X202102_divvy_tripdata,X202103_divvy_tripdata,X202104_divvy_tripdata,X202105_divvy_tripdata,X202106_divvy_tripdata,X202107_divvy_tripdata,X202108_divvy_tripdata,X202109_divvy_tripdata,X202110_divvy_tripdata,X202111_divvy_tripdata,X202112_divvy_tripdata)
```

## Combining tables into one dataset.

```{r combine into one file}
trips2021 <- bind_rows(X202101_divvy_tripdata,X202102_divvy_tripdata,X202103_divvy_tripdata,X202104_divvy_tripdata,X202105_divvy_tripdata,X202106_divvy_tripdata,X202107_divvy_tripdata,X202108_divvy_tripdata,X202109_divvy_tripdata,X202110_divvy_tripdata,X202110_divvy_tripdata,X202111_divvy_tripdata,X202112_divvy_tripdata)
```

## Analyzing new data set.

Now let's check out our results, it will show a huge file with all the data for entire 2021 year.

```{r view}
view(trips2021)
```
Use Summary function to get even more insights into data
```{r summary}
summary(trips2021)
```

## Creating additional columns as as month, date and day of the week.

For analysis by month, date, and day of the week, extract separately from the started_at column. 
We first extract dates only from the started_at column to create new date columns, and then extract months, day, and day of the week.

These extra columns will help us calculate trip length in our next step.

```{r Create columns as month, day, and day of the week}
trips2021$date <- as.Date(trips2021$started_at)
trips2021$month <- format(as.Date(trips2021$date), "%m")
trips2021$day <- format(as.Date(trips2021$date), "%d")
trips2021$year <- format(as.Date(trips2021$date), "%Y")
trips2021$day_of_the_week <- format(as.Date(trips2021$date), "%A")
```

## Calculating length of each trip and changing data format.

We will subtract values of started_at from the value of ended_at to find the length of each trip, and create a ride_length column.

```{r ride length}
trips2021$ride_length <- difftime(trips2021$ended_at,trips2021$started_at)
```

Change ride_length data format to 'numeric'.
We need to do it for analysis and calculations.

```{r ride_length to numeric}
trips2021$ride_length <- as.numeric(as.character(trips2021$ride_length))
```

#Let's check out our new columns.

```{r view}
str(trips2021)
```

## Remove rides that are less than zero.

Attach a version number to distinguish it. Add version 1.

```{r Remove unnecessary values}
 trips2021_final <- trips2021[!(trips2021$ride_length<0),]
```

This removed 147 observations from our original data frame.

## Final inspection before extracting file to creating visuals in Tableau.

```{r summary}
 summary(trips2021_final)
```

##File Extraction.

Now that we are done with data preparation we can extract csv file to import it to Tableau.

```{r File extraction}
write.csv(trips2021_final, "trips2021.csv")
```

































































